% =============================================================
% Lecture: Panel Data Econometrics
% Topic: Lecture 1 Quiz
% Course: Quantitative Economics, Vilnius University
% Author: Swapnil Singh
% =============================================================
\documentclass[10pt]{beamer}

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{appendixnumberbeamer}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{mathtools}
\usepackage{hyperref}

\usetheme{SimplePlus}

\title{Panel Data Econometrics}
\subtitle{Lecture 1 Quiz: Foundations}
\author{Swapnil Singh}
\institute{Lietuvos Bankas \and Kaunas University of Technology}
\date{Vilnius University, 2026}

\DeclareMathOperator{\Cov}{Cov}
\DeclareMathOperator{\Var}{Var}
\newcommand{\E}{\mathbb{E}}
\newcommand{\plim}{\operatorname*{plim}}

\begin{document}

\begin{frame}[plain]
  \titlepage
  \vfill
  {\scriptsize\color{gray} Based on \texttt{lectures/lecture\_panel\_data\_1.tex}.}
\end{frame}

\begin{frame}{How to use this quiz}
\small
\begin{block}{Coverage}
These five questions cover the core logic of Lecture 1:
\begin{itemize}
  \item why panel data change the comparison,
  \item why pooled OLS often fails,
  \item what fixed effects identify,
  \item how first differences compare with fixed effects,
  \item what two-way fixed effects absorb.
\end{itemize}
\end{block}

\begin{alertblock}{Deliberate omission}
Inference and clustering are excluded here. This quiz is about identification, interpretation, and algebra.
\end{alertblock}
\end{frame}

\begin{frame}{Quiz 1}
\small
\begin{block}{Question}
Suppose we study the effect of worker training $Train_{it}$ on wages $Wage_{it}$ using panel data.

Why is the panel comparison often more credible than the pure cross-sectional comparison? State the key source of omitted-variable bias that panel data may help remove.
\end{block}
\end{frame}

\begin{frame}{Solution 1}
\small
\begin{exampleblock}{Answer}
The cross-sectional comparison mixes workers with different permanent traits, such as ability, motivation, or firm quality. If those traits affect both training and wages, then
\[
\Cov(Train_{it},\alpha_i)\neq 0,
\]
where $\alpha_i$ is a time-invariant worker effect.
\end{exampleblock}

\begin{block}{Interpretation}
Panel data let us compare a worker to the worker's own past. If the omitted trait is stable over time, fixed effects or first differences can remove $\alpha_i$ and shift attention from across-worker comparisons to within-worker changes.
\end{block}
\end{frame}

\begin{frame}{Quiz 2}
\small
\begin{block}{Question}
Consider the scalar panel model
\[
y_{it} = \beta x_{it} + \alpha_i + u_{it}.
\]
Show the probability limit of pooled OLS and explain in one sentence why pooled OLS can be inconsistent even when $\Cov(x_{it},u_{it})=0$.
\end{block}
\end{frame}

\begin{frame}{Solution 2}
\small
\begin{exampleblock}{Answer}
The pooled OLS probability limit is
\[
\plim \hat{\beta}^{POLS}
=
\beta
+ \frac{\Cov(x_{it},\alpha_i)}{\Var(x_{it})}
+ \frac{\Cov(x_{it},u_{it})}{\Var(x_{it})}.
\]
\end{exampleblock}

\begin{block}{Interpretation}
Even if $x_{it}$ is uncorrelated with the idiosyncratic shock $u_{it}$, pooled OLS is still inconsistent whenever $x_{it}$ is correlated with the permanent unit effect $\alpha_i$. Then pooled OLS uses contaminated between-unit variation.
\end{block}
\end{frame}

\begin{frame}{Quiz 3}
\small
\begin{block}{Question}
Start from
\[
y_{it}=x_{it}'\beta+\alpha_i+u_{it}.
\]
Write the within-transformed equation. Then answer: why can fixed effects not estimate the coefficient on a regressor that is constant over time within each unit?
\end{block}
\end{frame}

\begin{frame}{Solution 3}
\small
\begin{exampleblock}{Within transformation}
Subtract the unit mean:
\[
y_{it}-\bar y_i = (x_{it}-\bar x_i)'\beta + (u_{it}-\bar u_i).
\]
Equivalently,
\[
\ddot y_{it}=\ddot x_{it}'\beta+\ddot u_{it}.
\]
\end{exampleblock}

\begin{block}{Why time-invariant regressors drop out}
If a regressor does not vary within unit, then $x_{it}=\bar x_i$ for all $t$, so
\[
x_{it}-\bar x_i=0.
\]
Fixed effects therefore have no within-unit variation left from which to identify its coefficient.
\end{block}
\end{frame}

\begin{frame}{Quiz 4}
\small
\begin{block}{Question}
In the model
\[
y_{it}=x_{it}\beta+\alpha_i+u_{it},
\]
compare fixed effects and first differences.

\begin{enumerate}
  \item When do they produce exactly the same slope coefficient?
  \item Why can first differencing be especially sensitive to measurement error?
\end{enumerate}
\end{block}
\end{frame}

\begin{frame}{Solution 4}
\small
\begin{exampleblock}{Part 1}
Fixed effects and first differences coincide when $T=2$. With only two periods, demeaning and differencing extract the same single within-unit movement.
\end{exampleblock}

\begin{exampleblock}{Part 2}
If the observed regressor is
\[
x_{it}^{obs}=x_{it}^\ast+\nu_{it},
\]
then differencing gives
\[
\Delta x_{it}^{obs}=\Delta x_{it}^\ast+(\nu_{it}-\nu_{i,t-1}).
\]
So the transformed regressor contains two measurement-error draws, which can magnify attenuation bias.
\end{exampleblock}
\end{frame}

\begin{frame}{Quiz 5}
\small
\begin{block}{Question}
Consider the two-way fixed-effects model
\[
y_{it}=x_{it}'\beta+\alpha_i+\lambda_t+u_{it}.
\]

Suppose one regressor can be written as
\[
x_{it}=a_i+b_t.
\]
Can its coefficient be identified under two-way fixed effects? Explain briefly.
\end{block}
\end{frame}

\begin{frame}{Solution 5}
\small
\begin{exampleblock}{Answer}
No. Under two-way fixed effects, the identifying variation must survive both the unit sweep and the time sweep. But a regressor of the form
\[
x_{it}=a_i+b_t
\]
is completely absorbed by the unit effects and time effects.
\end{exampleblock}

\begin{block}{Double-demeaning logic}
After the transformation
\[
\tilde x_{it}=x_{it}-\bar x_i-\bar x_t+\bar x,
\]
we get $\tilde x_{it}=0$ for every $(i,t)$. So there is no residual variation left to estimate its coefficient.
\end{block}
\end{frame}

\begin{frame}{Takeaways}
\small
\begin{enumerate}
  \item Panel data improve credibility by shifting the comparison toward within-unit changes.
  \item Pooled OLS often fails because it mixes within variation with contaminated between variation.
  \item Fixed effects identify slopes from within-unit movements only.
  \item First differences remove the same unit effect, but behave differently when $T>2$ and can amplify measurement error.
  \item Two-way fixed effects absorb both permanent unit differences and common period shocks.
\end{enumerate}
\end{frame}

\end{document}
