---
title: "Endogeneity vs Exogeneity: Why OLS Sometimes Fails"
author: "Swapnil Singh"
subtitle: "Quantitative Economics"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
    fig_width: 10
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

source("endogeneity_lab.R")
set_plot_style()
options(digits = 3)
```

# Learning goals

1. Understand the regression model $Y_i = \beta_0 + \beta_1 X_i + u_i$
2. See why OLS needs $\operatorname{cov}(X_i, u_i) = 0$
3. See what changes when $X$ becomes correlated with the hidden error term $u$
4. Build intuition for the three standard sources of endogeneity that we will study next

---

# The central idea

In a simple regression, $u_i$ collects **everything that affects $Y_i$ but is not written explicitly in the model**.

If $X_i$ is unrelated to that hidden part, OLS has a chance:
$$
\operatorname{cov}(X_i, u_i) = 0.
$$

If $X_i$ is related to the hidden part, OLS mixes up two forces:

- the genuine causal effect of $X$ on $Y$
- the hidden information inside $u$

That is **endogeneity**:
$$
\operatorname{cov}(X_i, u_i) \neq 0.
$$

---

# Step 1. A clean exogenous world

In the next cell, `gamma = 0.0`. That means the regressor `X` is not built from the hidden error `u`.

```{r exogenous}
beta_true <- 1.0

exogenous <- sample_basic_endogeneity(
  n     = 300,
  beta  = beta_true,
  gamma = 0.0,
  seed  = 2026
)

knitr::kable(basic_case_summary(exogenous, beta_true), digits = 3)
plot_basic_case(exogenous, beta_true, "Exogenous case")
```

In the summary above, focus on two numbers:

- `corr(X, u)`: it should be close to zero
- `OLS slope`: it should be close to the true slope $\beta = 1$

In the right-hand figure we can literally see the hidden error $u$ because this is a simulation. In real data we do **not** observe $u$, which is exactly why endogeneity is hard.

---

# Step 2. Now make the regressor endogenous

In the next cell, `gamma = 0.9`. That means part of `X` is mechanically built from `u`.

```{r endogenous}
endogenous <- sample_basic_endogeneity(
  n     = 300,
  beta  = beta_true,
  gamma = 0.9,
  seed  = 2026
)

knitr::kable(basic_case_summary(endogenous, beta_true), digits = 3)
plot_basic_case(endogenous, beta_true, "Endogenous case")
```

The pictures should now look different:

- `corr(X, u)` is no longer near zero
- the OLS line moves away from the true line

This is the whole problem in one sentence: **OLS is biased because the regressor carries hidden information from the error term.**

---

# Step 3. Monte Carlo: bias grows as endogeneity gets stronger

One sample can be unlucky. So we repeat the same experiment many times and average the results.

```{r monte-carlo-basic}
gamma_grid <- c(-1.0, -0.6, -0.3, 0.0, 0.3, 0.6, 1.0)

mc_results <- mc_basic_endogeneity(
  n            = 300,
  beta         = beta_true,
  gamma_values = gamma_grid,
  reps         = 600,
  seed         = 2026
)

knitr::kable(mc_results, digits = 3)
plot_basic_monte_carlo(mc_results, beta_true)
```

Read the table and figures as follows:

- when $\gamma = 0$, the average OLS estimate is close to the truth
- when $\gamma > 0$, OLS is biased upward
- when $\gamma < 0$, OLS is biased downward

The sign of the bias depends on the sign of the correlation between $X$ and $u$.

---

# What creates endogeneity in practice?

The algebra is always the same: $\operatorname{cov}(X, u) \neq 0$.

But the **mechanism** can differ. In introductory econometrics, the three standard mechanisms are:

1. **Omitted variable bias**: something important is left out of the regression, ends up inside $u$, and is correlated with $X$
2. **Measurement error**: the regressor is measured with noise, so the noise leaks into both $X$ and $u$
3. **Simultaneity**: $X$ and $Y$ are jointly determined, so $X$ already reflects shocks hitting $Y$

Notebook 2 studies those three cases one by one.

---

# Try it yourself

Change `trial_gamma`, re-knit (or run the chunk), and watch how the summary table and the figures move.

```{r try-it-yourself}
trial_gamma <- 0.4

trial <- sample_basic_endogeneity(
  n     = 300,
  beta  = beta_true,
  gamma = trial_gamma,
  seed  = 99
)

knitr::kable(basic_case_summary(trial, beta_true), digits = 3)
plot_basic_case(trial, beta_true,
                paste0("Trial case with gamma = ", trial_gamma))
```

---

# Takeaway

If $X$ is clean, OLS can recover the true slope. If $X$ is contaminated by the hidden part of the outcome equation, OLS loses its causal interpretation.

The next notebook answers the natural follow-up question: **what are the concrete economic stories that make $\operatorname{cov}(X, u) \neq 0$?**
